---
title: Increased fecundity associates with earlier offspring dispersal and son-bias in cooperative breeding Seychelles warblers
author: "Kiran Gok Lune Lee"
date: "2021"
geometry: margin=2cm 
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
    fig_width: 12
    fig_height: 10
  pdf_document:
    toc: yes
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)


```
## Load packages and data and create dataframes


```{r eval = TRUE, echo = TRUE}
setwd("~/Desktop/Seychelles Warblers/MSc_thesis_Kiran_Lee_2021")
library(ggplot2)
library(dplyr)
library(plyr)
library(tidyverse)
library(ggpubr)
library(lme4)
library(readxl)

# territory quality data
YearlyTQData <- read_excel("YearlyTQ.xlsx")
TQ <- read_excel("TQ.xlsx")
TQ<-rename(TQ, c("PeriodYear"="Year"))
TQ<-rename(TQ, c("TQcorrected"="TerritoryQuality"))
TQ<-subset(TQ,  !is.na(TerritoryQuality))
TQ$logTerritoryQuality<-log(TQ$TerritoryQuality)
TQ$GelmanTerritoryQuality<-(TQ$logTerritoryQuality)-(mean(TQ$logTerritoryQuality))/(2*sd(TQ$logTerritoryQuality))  # transform territory quality

#create dataframe with summary stats for each year on territory quality
TQgroupstata<-ddply(TQ, .(Year),summarise,
        MeanTQ= mean(TerritoryQuality),
        RangeTQ=max(TerritoryQuality)-min(TerritoryQuality),
        sdTQ= sd(TerritoryQuality),
        cvTQ= sdTQ/MeanTQ)

# sex ratio data
FullestSexRatioData <- read_excel("FullestSexRatioData.xlsx")
Data <- subset(FullestSexRatioData, !is.na(Year))
Data <- subset(Data, !is.na(Sex))
Data <- subset(Data, !is.na(TerritoryQuality))
Data <- subset(Data, pParentage>0.5) #choose chicks who's parentage is known with higher than half probability
Data$MumID[Data$MumID == "-998"] <- NA
Data$GelmanTerritoryQuality<-(log(Data$TerritoryQuality)-(mean(log(Data$TerritoryQuality))))/(2*sd(log(Data$TerritoryQuality))) # transform territory quality
ggplot(data=Data, aes(x=GelmanTerritoryQuality))+
  geom_histogram() #histogram to see spread of gelman territory quality. looks good


# status of each bird in field season
BirdIDSexYear <- read_excel("BirdIDSexYear.xlsx")

#create dataframe with summary stats per territory and then add this to other dataframes
TerritoryIDGroupStat<-ddply(Data, .(TerritoryID),summarise,
                   MeanTQbetweenyears= mean(TerritoryQuality),
                   RangeTQbetweenyears=max(TerritoryQuality)-min(TerritoryQuality),
                   sdTQbetweenyears= sd(TerritoryQuality),
                   SexRatiobetweenyears= mean(Sex),
                   cvTQbetweenyears= sdTQbetweenyears/MeanTQbetweenyears)
#add cvTQbetweenyears column
Data<-merge(x = Data , y = TerritoryIDGroupStat[ , c("TerritoryID", "cvTQbetweenyears")], by = "TerritoryID", all.x=TRUE)

# subset data so each row is a breedgroup in a field season
TData<-ddply(Data, .(Year, BreedGroupID, TerritoryID, nHelpers, nAB, nABX, nAllAB,nHelpersABX, nAdults,cvTQbetweenyears,nOffspringBreedGroup, FieldPeriodID),summarise, #This subset is important to not pseudoreplicate females offspring sex ratios. It takes the mean sex ratio when clutch size is bigger than 1.
             SexRatio= mean(Sex),
             TerritoryQuality= mean(TerritoryQuality),
             GelmanTerritoryQuality=mean(GelmanTerritoryQuality),
             nFem= length(Sex[Sex=="0"]), # number females per Year and territory ID
             nMale= length(Sex[Sex=="1"]),# number of males per year and territory ID
             nMums=n_distinct(MumID, na.rm = FALSE)
             
) 
#add column of how variable territory qualities are within a year
TData<-merge(x = TData , y = TQgroupstata[ , c("Year", "cvTQ")], by = "Year", all.x=TRUE)
#make breedgroupid a factor
TData$BreedGroupID<- as.factor(TData$BreedGroupID)
#add a column for subordinate non-helpers
TData$NonHelpers<-TData$nAdults - TData$nHelpersABX
#add in the dominant mum and dad id for each breed group
DominantMaleIDSexYear<-subset(BirdIDSexYear, Status=="BrM")
DominantMaleIDSexYear$MaleID<-DominantMaleIDSexYear$BirdID

DominantFemaleIDSexYear<-subset(BirdIDSexYear, Status=="BrF")
DominantFemaleIDSexYear$FemaleID<-DominantFemaleIDSexYear$BirdID

TData<-merge(x = TData, y = DominantMaleIDSexYear[ , c("BreedGroupID", "MaleID")], by = "BreedGroupID", all.x=TRUE)
TData<-merge(x = TData, y = DominantFemaleIDSexYear[ , c("BreedGroupID", "FemaleID")], by = "BreedGroupID", all.x=TRUE)
#add column of how variable a territory's quality is
TData<-merge(x = TData , y = TerritoryIDGroupStat[ , c("TerritoryID", "cvTQbetweenyears")], by = "TerritoryID", all.x=TRUE)
#add column of how variable territory qualities are within a year
TData<-merge(x = TData , y = TQgroupstata[ , c("Year", "cvTQ")], by = "Year", all.x=TRUE)
TData$BreedGroupID<- as.factor(TData$BreedGroupID)


##create dataframe with summary stats for each year on numbers daughters and males and territory quality 
DataGroupStat<-ddply(Data, .(Year),summarise,
                   MeanTQ= mean(TerritoryQuality),
                   RangeTQ=max(TerritoryQuality)-min(TerritoryQuality),
                   sdTQ= sd(TerritoryQuality),
                   OffspringSexRatio= mean(Sex), # mean offspring sex ratio had in that year
                   cvTQ= sdTQ/MeanTQ,
                   nDaughters= length(Sex[Sex=="0"]), # number daughters had in that year
                   nSons= length(Sex[Sex=="1"])) # number sons had in that year




# add fieldperiod start and end dates into dataframes
#first create dataframe with summary stats per fieldperiod
FieldPeriodIDGroupStat<-ddply(BirdIDSexYear, .(FieldPeriodID, PeriodEnd, PeriodStart),summarise,
                              nFem= length(Sex[Sex=="0"]),
                              nMale= length(Sex[Sex=="1"]))
FieldPeriodIDGroupStat$PeriodLength<-FieldPeriodIDGroupStat$PeriodEnd-FieldPeriodIDGroupStat$PeriodStart

Data<-merge(x = Data , y = FieldPeriodIDGroupStat [ , c("FieldPeriodID", "PeriodStart")], by = "FieldPeriodID", all.x=TRUE)
Data<-merge(x = Data , y = FieldPeriodIDGroupStat [ , c("FieldPeriodID", "PeriodEnd")], by = "FieldPeriodID", all.x=TRUE)
Data<-merge(x = Data , y = FieldPeriodIDGroupStat [ , c("FieldPeriodID", "PeriodLength")], by = "FieldPeriodID", all.x=TRUE)
TData<-merge(x = TData , y = FieldPeriodIDGroupStat [ , c("FieldPeriodID", "PeriodLength")], by = "FieldPeriodID", all.x=TRUE)

# create sex ratio data including mum's status
PedigreePlusStatus<- read_excel("PedigreePlusStatus.xlsx")
MumStatus <- subset(PedigreePlusStatus, !is.na(Year))
MumStatus <- subset(MumStatus, !is.na(Sex))
MumStatus <- subset(MumStatus, !is.na(TerritoryQuality))
MumStatus <- subset(MumStatus, pParentage>0.5) #choose chicks who's parentage is known with higher than half probability
MumStatus$MumID <- as.character(MumStatus$MumID)
MumStatus$MumID[MumStatus$MumID == "-998"] <- "NA"
MumStatus$MumID<- as.factor(MumStatus$MumID)
MumStatus$TerritoryQuality<-as.numeric(gsub(",", ".", MumStatus$TerritoryQuality))
MumStatus$GelmanTerritoryQuality<-(log(MumStatus$TerritoryQuality)-(mean(log(MumStatus$TerritoryQuality))))/(2*sd(log(MumStatus$TerritoryQuality)))
#add cvTQbetweenyears column
MumStatus<-merge(x = MumStatus , y = TerritoryIDGroupStat[ , c("TerritoryID", "cvTQbetweenyears")], by = "TerritoryID", all.x=TRUE)
Mumstatusgroupstat<-ddply(MumStatus, .(MumID,BreedGroupID,TerritoryID,GelmanTerritoryQuality,FieldPeriodID,nHelpers, nAB, nABX, nAllAB, nHelpersABX, nAdults,Year,cvTQbetweenyears,Status),summarise,
                    OffspringSexRatio= mean(Sex),
                    nFem= length(Sex[Sex=="0"]), # number females per treatment group
                    nMale= length(Sex[Sex=="1"]))

Mumstatusgroupstat$ClutchSize<-Mumstatusgroupstat$nFem+Mumstatusgroupstat$nMale
Mumstatusgroupstat<-merge(x = Mumstatusgroupstat , y = FieldPeriodIDGroupStat [ , c("FieldPeriodID", "PeriodLength")], by = "FieldPeriodID", all.x=TRUE)
# assign dominancy to mums
MumStatus$StatusBrF<-revalue(MumStatus$Status, c(ABX="NonBrF",H="NonBrF",AB="NonBrF",NSA="NonBrF",SEEN1="NonBrF",SEEN2="NonBrF",NS="NonBrF",OFL="NonBrF",SBR="NonBrF",TBRF="NonBrF"))
Mumstatusgroupstat$Dominant<-Mumstatusgroupstat$Status
Mumstatusgroupstat$Dominant[Mumstatusgroupstat$Status!="BrF"] <- 0
Mumstatusgroupstat$Dominant[Mumstatusgroupstat$Status=="BrF"] <- 1
Mumstatusgroupstat$NonHelpers<-Mumstatusgroupstat$nAdults - Mumstatusgroupstat$nHelpersABX
Mumstatusgroupstat$Dominant<-as.factor(Mumstatusgroupstat$Dominant)

#mum group stat. My dataset contains 995 mums and 1034 dads.
Mumgroupstat<-ddply(Data, .(MumID,BreedGroupID,GelmanTerritoryQuality,nHelpers, nAB, nABX, nAllAB,nHelpersABX, nAdults,Year,cvTQbetweenyears,PeriodLength),summarise,
                          OffspringSexRatio= mean(Sex),
                          nFem= length(Sex[Sex=="0"]), # number females per treatment group
                          nMale= length(Sex[Sex=="1"]))

#how many unique breeding events? 396 (when not including -998 who are NA mums)
length(unique(unlist(Mumgroupstat[c("MumID")])))

Mumgroupstat$ClutchSize<-Mumgroupstat$nFem+Mumgroupstat$nMale

#creating dataframes to study dispersal over subsequent years
Data$YearLater<-Data$Year+1
Data$TwoYearsLater<-Data$Year+2
Data$ThreeYearsLater<-Data$Year+3

#merge status column in BirdIDSexYear if the Year matches YearLater and if BirdID matches ChickID. Here I am basically creating three dataframes where for each bird, there is data on its status after one year (TrialMergeData), two years(TrialMergeData2) or three years (TrialMergeData3)
TrialMergeData = merge(Data, BirdIDSexYear, by.x=c("YearLater", "ChickID"), by.y=c("Year", "BirdID"))
TrialMergeData2 = merge(Data, BirdIDSexYear, by.x=c("TwoYearsLater", "ChickID"), by.y=c("Year","BirdID"))
TrialMergeData3 = merge(Data, BirdIDSexYear, by.x=c("ThreeYearsLater", "ChickID"), by.y=c("Year","BirdID"))

TrialMergeData$MumID<-as.factor(TrialMergeData$MumID)
TrialMergeData2$MumID<-as.factor(TrialMergeData2$MumID)
TrialMergeData3$MumID<-as.factor(TrialMergeData3$MumID)
Mumgroupstat$MumID<-as.factor(Mumgroupstat$MumID)

TrialMergeData$BreedGroupID<-TrialMergeData$BreedGroupID.x
TrialMergeData$BreedGroupID<-as.factor(TrialMergeData$BreedGroupID)
TrialMergeData2$BreedGroupID<-TrialMergeData2$BreedGroupID.x
TrialMergeData2$BreedGroupID<-as.factor(TrialMergeData2$BreedGroupID)
TrialMergeData3$BreedGroupID<-TrialMergeData3$BreedGroupID.x
TrialMergeData3$BreedGroupID<-as.factor(TrialMergeData3$BreedGroupID)
Mumgroupstat$BreedGroupID<-as.factor(Mumgroupstat$BreedGroupID)


TrialMergeData<-left_join(TrialMergeData, Mumgroupstat, by=c('MumID'='MumID', 'Year'='Year', 'BreedGroupID'='BreedGroupID'))
TrialMergeData2<-left_join(TrialMergeData2, Mumgroupstat, by=c('MumID'='MumID', 'Year'='Year', 'BreedGroupID'='BreedGroupID'))
TrialMergeData3<-left_join(TrialMergeData3, Mumgroupstat, by=c('MumID'='MumID', 'Year'='Year', 'BreedGroupID'='BreedGroupID'))

#remove NAs 
TrialMergeData$MumID[TrialMergeData$MumID == "NA"] <- NA
TrialMergeData$DadID[TrialMergeData$DadID == -998] <- NA
TrialMergeData<-subset(TrialMergeData, !is.na(MumID))
# assign ranks (1-15) to individuals based on their status
TrialMergeData$RankedStatus<-recode(TrialMergeData$Status, "CH"=1,"BrF" = 1, "BrM" = 1, "H"=2, "OFL" = 3, "FL" = 4, "ABX" = 5, "AB" = 6, "FLOAT" = 7, "SEEN2" = 8, "SEEN1" = 9, "TBRF" = 10, "TBRM" = 10, "U" = 11, "SBR" = 12, "B"=13,"NSA"=14, "NS"=15)

#select chicks whose rank the next year is "highest" if the same chick occurs twice
TrialMergeData<-TrialMergeData %>%
  group_by(ChickID, BreedGroupID, FieldPeriodID.x, Sex.x, YearLater, nOffspringBreedGroup, nAdults.x) %>%
  slice(which.min(RankedStatus))



#remove NAs 
TrialMergeData2$MumID[TrialMergeData2$MumID == "NA"] <- NA
TrialMergeData2$DadID[TrialMergeData2$DadID == -998] <- NA
TrialMergeData2<-subset(TrialMergeData2, !is.na(MumID))
# assign ranks (1-15) to individuals based on their status
TrialMergeData2$RankedStatus<-recode(TrialMergeData2$Status, "CH"=1,"BrF" = 1, "BrM" = 1, "H"=2, "OFL" = 3, "FL" = 4, "ABX" = 5, "AB" = 6, "FLOAT" = 7, "SEEN2" = 8, "SEEN1" = 9, "TBRF" = 10, "TBRM" = 10, "U" = 11, "SBR" = 12, "B"=13,"NSA"=14, "NS"=15)

#select chicks whose rank in two years is "highest" if the same chick occurs twice
TrialMergeData2<-TrialMergeData2 %>%
  group_by(ChickID, BreedGroupID, FieldPeriodID.x, Sex.x, YearLater, nOffspringBreedGroup, nAdults.x) %>%
  slice(which.min(RankedStatus))


#remove NAs 
TrialMergeData3$MumID[TrialMergeData3$MumID == "NA"] <- NA
TrialMergeData3$DadID[TrialMergeData3$DadID == -998] <- NA
TrialMergeData3<-subset(TrialMergeData3, !is.na(MumID))
# assign ranks (1-15) to individuals based on their status
TrialMergeData3$RankedStatus<-recode(TrialMergeData3$Status, "CH"=1,"BrF" = 1, "BrM" = 1, "H"=2, "OFL" = 3, "FL" = 4, "ABX" = 5, "AB" = 6, "FLOAT" = 7, "SEEN2" = 8, "SEEN1" = 9, "TBRF" = 10, "TBRM" = 10, "U" = 11, "SBR" = 12, "B"=13,"NSA"=14, "NS"=15)

#select chicks whose rank in three years is "highest" if the same chick occurs twice
TrialMergeData3<-TrialMergeData3 %>%
  group_by(ChickID, BreedGroupID, FieldPeriodID.x, Sex.x, YearLater, nOffspringBreedGroup, nAdults.x) %>%
  slice(which.min(RankedStatus))

#add a column for whether they remain on territory or not
TrialMergeData$DispersalStatus<-if_else( TrialMergeData$TerritoryID.x==TrialMergeData$TerritoryID.y, 0, 1)
TrialMergeData2$DispersalStatus<-if_else( TrialMergeData2$TerritoryID.x==TrialMergeData2$TerritoryID.y, 0, 1)
TrialMergeData3$DispersalStatus<-if_else( TrialMergeData3$TerritoryID.x==TrialMergeData3$TerritoryID.y, 0, 1)

#add columns to each dataframe specifying the number of years later for the data of the chick
TrialMergeData$YearsLater<-(TrialMergeData$YearLater/TrialMergeData$YearLater)
TrialMergeData2$YearsLater<-(TrialMergeData2$YearLater/TrialMergeData2$YearLater)+1
TrialMergeData3$YearsLater<-(TrialMergeData3$YearLater/TrialMergeData3$YearLater)+2
TrialMergeData$DispersalStatusOneYear<-TrialMergeData$DispersalStatus
TrialMergeData2$DispersalStatusTwoYears<-TrialMergeData2$DispersalStatus
TrialMergeData3$DispersalStatusThreeYears<-TrialMergeData3$DispersalStatus
TrialMergeData$BreedingStatusOneYear<-TrialMergeData$Status
TrialMergeData2$BreedingStatusTwoYears<-TrialMergeData2$Status
TrialMergeData3$BreedingStatusThreeYears<-TrialMergeData3$Status

# merging relevant data from each TrialMergeData into one dataframe where $YearsLater identifies the number of years since the data is based on
OneYearLater<-data.frame(TrialMergeData$ChickID,TrialMergeData$MumID,TrialMergeData$DadID, TrialMergeData$Sex.x, TrialMergeData$Status,TrialMergeData$DispersalStatus, TrialMergeData$YearsLater, TrialMergeData$nOffspringBreedGroup, TrialMergeData$ClutchSize,TrialMergeData$nAllAB.x,TrialMergeData$nAdults.x,TrialMergeData$BreedGroupID,TrialMergeData$TerritoryID.x,TrialMergeData$GelmanTerritoryQuality.x,TrialMergeData$TerritoryID.y,TrialMergeData$FieldPeriodID.y,TrialMergeData$FieldPeriodID.x,TrialMergeData$YearLater)
TwoYearLater<-data.frame(TrialMergeData2$ChickID,TrialMergeData2$MumID,TrialMergeData2$DadID, TrialMergeData2$Sex.x, TrialMergeData2$Status,TrialMergeData2$DispersalStatus, TrialMergeData2$YearsLater, TrialMergeData2$nOffspringBreedGroup, TrialMergeData2$ClutchSize, TrialMergeData2$nAllAB.x,TrialMergeData2$nAdults.x,TrialMergeData2$BreedGroupID,TrialMergeData2$TerritoryID.x,TrialMergeData2$GelmanTerritoryQuality.x,TrialMergeData2$TerritoryID.y,TrialMergeData2$FieldPeriodID.y,TrialMergeData2$FieldPeriodID.x,TrialMergeData2$YearLater)
ThreeYearLater<-data.frame(TrialMergeData3$ChickID,TrialMergeData3$MumID,TrialMergeData3$DadID, TrialMergeData3$Sex.x, TrialMergeData3$Status,TrialMergeData3$DispersalStatus, TrialMergeData3$YearsLater, TrialMergeData3$nOffspringBreedGroup, TrialMergeData3$ClutchSize,TrialMergeData3$nAllAB.x, TrialMergeData3$nAdults.x,TrialMergeData3$BreedGroupID,TrialMergeData3$TerritoryID.x,TrialMergeData3$GelmanTerritoryQuality.x,TrialMergeData3$TerritoryID.y,TrialMergeData3$FieldPeriodID.y,TrialMergeData3$FieldPeriodID.x,TrialMergeData3$YearLater)

#tidy up the spaces and fullstops in the column headings
names(OneYearLater) <- gsub(x = names(OneYearLater),
                            pattern = "\\.",
                            replacement = " ")
names(TwoYearLater) <- gsub(x = names(TwoYearLater),
                            pattern = "\\.",
                            replacement = " ")
names(ThreeYearLater) <- gsub(x = names(ThreeYearLater),
                            pattern = "\\.",
                            replacement = " ")

OneYearLater<-rename(OneYearLater, c("TrialMergeData ChickID"="ChickID",
                                     "TrialMergeData MumID"="MumID",
                                     "TrialMergeData DadID"="DadID",
                                     "TrialMergeData Sex x"="Sex",
                     "TrialMergeData Status"="BreedingStatus",
                     "TrialMergeData DispersalStatus"="DispersalStatus",
                     "TrialMergeData YearsLater"="YearsLater",
                     "TrialMergeData nOffspringBreedGroup"="nOffspringBreedGroup",
                     "TrialMergeData ClutchSize"="ClutchSize",
                     "TrialMergeData nAllAB x"="nAllABx",
                     "TrialMergeData nAdults x"="nAdults",
                     "TrialMergeData BreedGroupID"="BreedGroupID",
                     "TrialMergeData TerritoryID x"="TerritoryIDx",
                     "TrialMergeData GelmanTerritoryQuality x"="GelmanTerritoryQuality",
                     "TrialMergeData TerritoryID y"="TerritoryIDy",
                     "TrialMergeData FieldPeriodID y"="FieldPeriody",
                     "TrialMergeData FieldPeriodID x"="FieldPeriodx",
                     "TrialMergeData YearLater"="YearLater"))
TwoYearLater<-rename(TwoYearLater, c("TrialMergeData2 ChickID"="ChickID",
                                     "TrialMergeData2 MumID"="MumID",
                                     "TrialMergeData2 DadID"="DadID",
                                     "TrialMergeData2 Sex x"="Sex",
                                     "TrialMergeData2 Status"="BreedingStatus",
                                     "TrialMergeData2 DispersalStatus"="DispersalStatus",
                                     "TrialMergeData2 YearsLater"="YearsLater",
                                     "TrialMergeData2 nOffspringBreedGroup"="nOffspringBreedGroup",
                                     "TrialMergeData2 ClutchSize"="ClutchSize",
                                     "TrialMergeData2 BreedGroupID"="BreedGroupID",
                                     "TrialMergeData2 TerritoryID x"="TerritoryIDx",
                                     "TrialMergeData2 GelmanTerritoryQuality x"="GelmanTerritoryQuality",
                                     "TrialMergeData2 TerritoryID y"="TerritoryIDy",
                                     "TrialMergeData2 FieldPeriodID y"="FieldPeriody",
                                     "TrialMergeData2 FieldPeriodID x"="FieldPeriodx",
                                     "TrialMergeData2 nAllAB x"="nAllABx",
                                     "TrialMergeData2 nAdults x"="nAdults",
                                     "TrialMergeData2 YearLater"="YearLater"))
ThreeYearLater<-rename(ThreeYearLater, c("TrialMergeData3 ChickID"="ChickID",
                                         "TrialMergeData3 MumID"="MumID",
                                         "TrialMergeData3 DadID"="DadID",
                                     "TrialMergeData3 Sex x"="Sex",
                                     "TrialMergeData3 Status"="BreedingStatus",
                                     "TrialMergeData3 DispersalStatus"="DispersalStatus",
                                     "TrialMergeData3 YearsLater"="YearsLater",
                                     "TrialMergeData3 nOffspringBreedGroup"="nOffspringBreedGroup",
                                     "TrialMergeData3 ClutchSize"="ClutchSize",
                                     "TrialMergeData3 BreedGroupID"="BreedGroupID",
                                     "TrialMergeData3 TerritoryID x"="TerritoryIDx",
                                     "TrialMergeData3 GelmanTerritoryQuality x"="GelmanTerritoryQuality",
                                     "TrialMergeData3 TerritoryID y"="TerritoryIDy",
                                     "TrialMergeData3 FieldPeriodID y"="FieldPeriody",
                                     "TrialMergeData3 FieldPeriodID x"="FieldPeriodx",
                                     "TrialMergeData3 nAllAB x"="nAllABx",
                                     "TrialMergeData3 nAdults x"="nAdults",
                                     "TrialMergeData3 YearLater"="YearLater"))

AllYearLater<-bind_rows(OneYearLater, TwoYearLater, ThreeYearLater)
#rank status to numbers so that we can exclude those not seen and not known to be alive
AllYearLater$RankedStatus<-AllYearLater$BreedingStatus
AllYearLater$RankedStatus<-recode(AllYearLater$RankedStatus, "CH"=1,"BrF" = 1, "BrM" = 1, "H"=2, "OFL" = 3, "FL" = 4, "ABX" = 5, "AB" = 6, "FLOAT" = 7, "SEEN2" = 8, "SEEN1" = 9, "TBRF" = 10, "TBRM" = 10, "U" = 11, "SBR" = 12, "B"=13,"NSA"=14, "NS"=15)
AllYearLater$BreedGroupSize<-AllYearLater$nAdults+AllYearLater$nOffspringBreedGroup
AllYearLater<-subset(AllYearLater, RankedStatus<14) #remove chicks listed as "NS (not seen) and NSA (not seen known to be alive)




```
## Offspring that could be assigned genetic parentage with at least 50% confidence (mean +/- SE confidence of parentage = 0.98 +/- 0.002, n= 1140) were included
```{r eval = TRUE, echo = TRUE}
mean(Data$pParentage)
sd(Data$pParentage, na.rm=TRUE) /  
  sqrt(length(Data$pParentage[!is.na(Data$pParentage)]))
nrow(Data)
```
## three subsequent years as most offspring (this study: 92.1%, n=508) disperse by then
```{r eval = TRUE, echo = TRUE}
table(AllYearLater$YearsLater, AllYearLater$DispersalStatus)

```
## Most mothers have singleton clutches per breeding season, though 14.7% (144/977) produce multiple offspring, up to four per breeding season (mean +/- SE= 1.17 +/- 0.014). 
```{r eval = TRUE, echo = TRUE}
table(Mumgroupstat$ClutchSize)
mean(Mumgroupstat$ClutchSize)
sd(Mumgroupstat$ClutchSize, na.rm=TRUE) /  
  sqrt(length(Mumgroupstat$ClutchSize[!is.na(Mumgroupstat$ClutchSize)]))
```
## 13.7% (128/931) of mothers are co-breeding subordinates NB: This statement in my report is incorrect if we run the script (126/932).
```{r eval = TRUE, echo = TRUE}
table(MumStatus$StatusBrF)
```
## the number of offspring produced by breed groups, with an upper limit of seven per breeding season (mean +/- SE= 1.68 +/- 0.031) 
NB: This statement in my report is incorrect if we run the script (1.35+/- 0.027). 
```{r eval = TRUE, echo = TRUE}
mean(TData$nOffspringBreedGroup)
sd(TData$nOffspringBreedGroup, na.rm=TRUE) /  
  sqrt(length(TData$nOffspringBreedGroup[!is.na(TData$nOffspringBreedGroup)]))
```

## Breed groups with more helping subordinates, but not non-helping subordinates produce more offspring (figure 1), with no impact of territory quality or field period length in which sampling took place (table 1)
```{r eval = TRUE, echo = TRUE}
#Territories with more helpers have larger clutches visually
table(TData$nHelpers)

ggplot(data=subset(TData), aes(x=factor(nHelpers), y=nOffspringBreedGroup))+
 geom_violin() +
  stat_summary(fun.data=mean_sdl, mult=1, 
                 geom="pointrange", color="red")+
  geom_jitter(alpha = .25, width = .2, height = 0.1) +
  ylim(1,7)+ 
  labs(x= "number of helping subordinates", y="number of offspring born")+
  theme(axis.title.x = element_text(size = rel(2)),
        axis.title.y = element_text(size = rel(2)),
        axis.text =element_text(size = rel(1.5)))+
  annotate("text", size=5,y = 0.5, x = 1, label = "n=656")+
  annotate("text",size=5, y = 0.5, x = 2, label = "n=180")+
  annotate("text", size=5, y = 0.5, x = 3, label = "n=25")+
  annotate("text", size=5,y = 0.5, x =4, label = "n=2")+
  ylim(0.5,7)

mm<-glmer(nOffspringBreedGroup ~ GelmanTerritoryQuality +nHelpers + NonHelpers +PeriodLength + (1|TerritoryID) + (1|FieldPeriodID)+ (1|MaleID)+ (1|FemaleID), data=TData, family=poisson)
summary(mm)
drop1(mm, test="Chisq")
```
## Breed groups in better quality territories sustain more adult individuals (estimate= 0.179, SE= 0.080, p= 0.025, figure 2)
```{r eval = TRUE, echo = TRUE}
mmm<-glmer(nAdults~GelmanTerritoryQuality + (1|FieldPeriodID) + (1|TerritoryID) + (1|MaleID) + (1|FemaleID), data=TData, family=poisson)
summary(mmm)
drop1(mmm, test="Chisq")

ggplot(data=TData, aes(x=factor(nAdults) , y=GelmanTerritoryQuality))+
  geom_violin() + coord_flip() +
  stat_summary(fun.y=mean, mult=1, 
               geom="pointrange", color="red")+
  geom_jitter(alpha = .25, width = .2, height = 0.1)+
  labs(x= "number of subordinates", y="territory quality")+
  theme(axis.title.x = element_text(size = rel(2)),
        axis.title.y = element_text(size = rel(2)),
        axis.text =element_text(size = rel(1.5)))+
  annotate("text", size=5,y = -1.9, x = 7, label = "n=6")+
  annotate("text",size=5, y = -1.9, x = 6, label = "n=10")+
  annotate("text", size=5, y = -1.9, x = 5, label = "n=32")+
  annotate("text", size=5,y = -1.9, x =4, label = "n=83")+
  annotate("text",size=5, y = -1.9, x = 3, label = "n=203")+
  annotate("text", size=5, y = -1.9, x = 2, label = "n=295")+
  annotate("text", size=5, y = -1.9, x = 1, label = "n=234")+ ylim(-2.01,2.01)
```

## Per mother, the number of offspring produced is not influenced by territory quality (estimate= 0.039, SE= 0.068, p= 0.566), by the number of helping subordinates (estimate=  0.0610, SE= 0.068, p= 0.566),  or by non-helping subordinates in the breed group (estimate= 0.003, SE= 0.035, p= 0.938), whether she is dominant or subordinate (estimate= 0.032, SE= 0.102, p= 0.756), or the field period length in which sampling took place (estimate<0.001, SE= 0.002, p= 0.893)
```{r eval = TRUE, echo = TRUE}
mmm<-glmer(ClutchSize ~ GelmanTerritoryQuality+nHelpers+Dominant +  NonHelpers+ PeriodLength + (1|TerritoryID) + (1|MumID), data=subset(Mumstatusgroupstat), family=poisson)

summary(mmm)
```

## 53.2% (453/851) of offspring disperse after one year, 85.9% (556/647) after two years and 92.1% (468/508) after three years

```{r eval = TRUE, echo = TRUE}
table(AllYearLater$YearsLater, AllYearLater$DispersalStatus)

```
## Rates of offspring dispersal from the natal territory increases when offspring natal territory quality is higher and when more offspring are born in the breed group (table 2, Fig. 2). 
This script no longer produces the trend line for Fig. 2 in my thesis. I have tried troubleshooting but without success.
```{r eval = TRUE, echo = TRUE}
m2<-glmer(DispersalStatus~Sex+GelmanTerritoryQuality+nOffspringBreedGroup*YearsLater+nAdults+(1|BreedGroupID)+(1|FieldPeriodx)+(1|TerritoryIDx)+(1|MumID)+(1|DadID), data=AllYearLater, family = binomial)
summary(m2)



AllYearLater$nOffspringBreedGroupMerged<-AllYearLater$nOffspringBreedGroup
AllYearLater$nOffspringBreedGroupMerged[AllYearLater$nOffspringBreedGroupMerged>2]<-"3+"

ggplot(data=subset(AllYearLater, YearsLater!=0), aes(x=factor(nOffspringBreedGroupMerged), y=(DispersalStatus), color=factor(YearsLater))) + 
  stat_summary(fun.data=mean_sdl, mult=1, 
               geom="point",line="dotted", size=3)+
  stat_summary(aes(group=1, color=factor(YearsLater)),fun.data=mean_sdl, mult=1, 
               geom="line")+
  geom_jitter(alpha=0.25,width=0.3,height=0.05)+ 
  geom_smooth(method='lm', se=F)+ ylim(-0.15,1) + 
  labs(x= "offspring born", y="dispersal status",color = "years later") +
  theme(legend.direction = "vertical")+
  guides(color = guide_legend(reverse=TRUE)) +
  theme(axis.title.x = element_text(size = rel(1.5)),
        axis.title.y = element_text(size = rel(1.5)),
        axis.text =element_text(size = rel(1)),
        legend.text = element_text(size = rel(1.5)),
        legend.title = element_text(size = rel(1.5)))+
  theme(legend.background=element_blank())+
  annotate("text", size=5,x = 1, y = -0.15, label = "n=1145")+
  annotate("text",size=5, x = 2, y = -0.15, label = "n=561")+
  annotate("text", size=5, x = 3, y = -0.15, label = "n=300")
  

```
## Sons become dominants more than daughters in each year (estimate= 0.301, SE= 0.101 p= 0.003) 


```{r eval = TRUE, echo = TRUE}
AllYearLater$Dominant<-AllYearLater$RankedStatus
AllYearLater$Dominant[AllYearLater$RankedStatus==1] <- 1
AllYearLater$Dominant[AllYearLater$RankedStatus>1] <- 0
m<-glmer(Dominant ~ Sex + GelmanTerritoryQuality+ nOffspringBreedGroup +nAllABx+ (1|TerritoryIDx) +(1|FieldPeriodx) + (1|MumID) +(1|DadID), data=AllYearLater, family=binomial)
summary(m)
```
## 89.6% (241/269) of sons becoming dominant and 82.8% (198/239) of daughters becoming dominant three years later


```{r eval = TRUE, echo = TRUE}
table(AllYearLater$YearsLater, AllYearLater$Dominant,AllYearLater$Sex)
241/(241+28) # sons
198/(198+41) # daughters
```
## There is no impact of natal territory quality (estimate= 0. 252, SE= 0.150, p= 0.093), the number of offspring co-born in the natal territory (estimate= -0.042, SE= 0.059, p= 0.476) or the number of adult subordinates in the natal territory (estimate= -0.061, SE= 0.069, p= 0.380) on whether offspring established dominancy in subsequent years.


```{r eval = TRUE, echo = TRUE}
summary(m)
```
## Dominant mothers male-bias offspring sex ratios when increasing the number of offspring produced, despite even mean offspring sex-ratios (0.48) (table 3, Fig. 3a)...There is no influence of territory quality or the number of adult subordinates in the territory on offspring sex-ratios


```{r eval = TRUE, echo = TRUE}
m<-glmer(cbind(nMale,nFem)~ClutchSize+Dominant+GelmanTerritoryQuality +nHelpers + cvTQbetweenyears + (1|MumID) + (1|FieldPeriodID) + (1|TerritoryID), data=subset(Mumstatusgroupstat, !is.na(cvTQbetweenyears)), family=binomial)
summary(m)

Mumstatusgroupstat$ClutchSizeMerged<-Mumstatusgroupstat$ClutchSize
Mumstatusgroupstat$ClutchSizeMerged[Mumstatusgroupstat$ClutchSizeMerged>2]<-"3+"
lables <- c("1" = "Dominant", "0" = "Subordinate")
dat_text <- data.frame(
  Dominant = c("1", "1", "1", "0","0","0"),
  lables = c( "n=577","n=95","n=13","n=93", "n=10", "n=4"),
  x     = c(1, 2,3,1,2,3),
  y     = c(-0.1, -0.1, -0.1,-0.1,-0.1,-0.1)
)
ggplot(subset(Mumstatusgroupstat), aes(x=ClutchSizeMerged, y=OffspringSexRatio,color=Dominant, group=Dominant)) +
  geom_jitter(width=0.4,height=0.01)+
  geom_smooth(method = lm, se=F) + 
  stat_summary(
    geom = "point",
    fun.y = "mean",
    col = "black",
    size = 3,
    shape = 21,
    fill = "black"
  ) + 
  facet_grid(col=vars(Dominant), labeller=labeller(Dominant = lables))+
  geom_hline(data=filter(Mumstatusgroupstat, Status=="BrF"), 
          aes(yintercept=mean(Mumstatusgroupstat$OffspringSexRatio[Mumstatusgroupstat$Status=="BrF"])), 
          linetype="dashed", 
          size=1, width=0.5, 
          colour="#00CCCC") + 
  geom_hline(data=filter(Mumstatusgroupstat, Status!="BrF"), 
          aes(yintercept=mean(Mumstatusgroupstat$OffspringSexRatio[Mumstatusgroupstat$Status!="BrF"])),
          linetype="dashed", 
          size=1, width=0.5,
          colour="#FF3333") + 
  labs(x = "offspring produced", y = "offspring sex ratio")+
  theme(axis.title.x = element_text(size = rel(2)),
        axis.title.y = element_text(size = rel(2)),
        axis.text =element_text(size = rel(1.5)),
        legend.title =element_text(size = rel(1.5)),
        legend.text =element_text(size = rel(1.5) ),
        strip.background = element_blank(),
        strip.text.x = element_blank())+ 
  scale_color_discrete(name = "mother status",breaks=c("0", "1"), labels = c("subordinate", "dominant"))+
  geom_text(size=5,data = dat_text, aes(x = x,  y = y, label = lables), show.legend = FALSE)
```
## maintaining a male-biased mean offspring sex-ratio of 0.62 (78/126, two-tailed binomial test= 0.009) 

```{r eval = TRUE, echo = TRUE}
binom.test(x=78, n=126, p = 1/2,
           alternative = "two.sided")
```
